---
title: "LASSO Regression - Spring 2026 "
subtitle: "Determining Survival Rates for Breast Cancer Patients"
author: "Stephanie Hopkins (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)

## Introduction

-   Clearly state the problem or question you're addressing.

-   Introduce why it is relevant needs.

-   Provide an overview of your approach.

The LASSO, Least Absolute Shrinkage and Selector Operator, regression
model has become very popular over the last 30 years. The model was
first introduced by Robert Tibshirani to perform variable selection and
regularization to improve prediction accuracy and interpretability
[@emmert2019high]. Over the years, it has been used to research topics
in a numerous amount of fields including biology, computational social
sciences, social media, the stock market, economy, political science,
robotics, climatology, and pharmacology [@emmert2019high]. In LASSO
regression, the prediction error is minimized by identifying the
variables and corresponding regression coefficients and then putting
constraints on the model which pushes the regression coefficients to
zero [@ranstam2018lasso]. Due to this, this regression model works
extremely well with outliers, thus dealing with large-scale problems
very efficiently [@roth2004generalized].

The goal of this analysis, using LASSO regression, is to estimate the
survival rate of breast cancer patients. ...

## Methods

-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.

LASSO regression can be defined by [@emmert2019high]:

$$
\hat{\beta} = arg \space \min \{\frac{1}{2n} \sum_{i=1}^{n} (y_i - \sum_{j} \beta_j x_{ij})^2 + \lambda||\beta||_1\} 
$$

where (define variables)

## Analysis and Results

### Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

A study was conducted to determine how...

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{python}
import pandas as pd
```

```{r, warning=FALSE, echo=TRUE}
# Load Data
kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

### Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
