---
title: "LASSO Regression - Spring 2026 "
subtitle: "Determining Predictors in Survival for Breast Cancer Patients"
author: "Stephanie Hopkins & Aminata Gado Oumarou (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)

## Introduction

The LASSO (Least Absolute Shrinkage and Selection Operator) regression
model has become increasingly popular over the past 30 years. It was
first introduced by Robert Tibshirani as a method for variable selection
and regularization to improve prediction accuracy and interpretability.
Since then, LASSO has been widely applied across many fields, including
biology, computational social sciences, social media, the stock market,
economics, political science, robotics, climatology, and pharmacology
[@emmert2019high].

Regression is a fundamental task in supervised machine learning, used to
model relationships between variables and address key challenges such as
parameter estimation and variable selection [@muthukrishnan2016lasso].
As machine learning methods continue to grow in popularity and show
strong results, their importance in research has increased. In medicine
especially, predictive models developed through machine learning can
help professionals distinguish between risk groups, design personalized
treatment plans, and improve patient survival outcomes
[@huang2023practical].

LASSO regression works by minimizing prediction error while applying a
constraint that shrinks some regression coefficients toward zero
[@ranstam2018lasso]. It does this by adding an L1 penalty term to the
loss function, which constrains the sum of the absolute values of the
coefficients [@ranstam2018lasso]. This penalty forces weaker predictors
to exactly zero, effectively performing automatic variable selection
[@usman2021comparing]. While traditional LASSO in linear regression
minimizes the residual sum of squares subject to an L1 constraint
[@tibshirani1996regression], the method can be extended to survival
analysis through the Cox proportional hazards model. In survival
analysis, the Cox proportional hazards model is widely used to examine
the relationship between covariates and time-to-event outcomes, such as
overall survival. Rather than minimizing residual error, the Cox model
estimates regression coefficients by maximizing the partial likelihood
function. The LASSO-Cox model incorporates an L1 penalty into this
partial likelihood, allowing some regression coefficients to shrink
exactly to zero while estimating hazard ratios [@tibshirani1997lasso].
This approach reduces multicollinearity, improves estimation stability,
and produces a sparse and interpretable model, which is especially
valuable in high-dimensional clinical datasets [@shahraki2015survival].

The goal of this analysis is to use LASSO-Cox regression to identify the
factors most strongly associated with breast cancer survival time.
Cancer continues to have a major impact worldwide. In 2025, an estimated
2,041,910 new cancer cases will be diagnosed in the United States, and
approximately 618,120 people are expected to die from the disease
[@nci2025statistics]. Breast cancer remains a significant public health
concern. By 2019, more than 3.8 million individuals in the United States
were living with breast cancer, and it is still the most commonly
diagnosed cancer among women and the second most common cancer globally
[@usman2021comparing]. Breast cancer begins in the breast’s tissues and
develops in the epithelial cells of the ducts (85% of cases) and lobules
(15%). Causes include hereditary factors, age, postmenopausal obesity,
alcohol or smoking, and other factors [@hadba2024comparison].
Identifying the clinical and demographic factors most strongly linked to
survival can help improve prognostic models and support better clinical
decision-making. Previous research has identified numerous factors
associated with breast cancer prognosis and survival. These factors are
commonly grouped into three categories: (1) demographic and genetic
characteristics, such as age at diagnosis, marital status, and
reproductive history; (2) clinicopathological factors, including tumor
location, size, and histologic grade; and (3) treatment-related
variables, such as drug therapy, radiotherapy, and chemotherapy
[@teng2022dynamic].

To accomplish this, the analysis uses the Surveillance, Epidemiology,
and End Results (SEER) dataset to determine the most important variables
associated with breast cancer survivability. SEER has collected and
maintained high-quality, validated data on mortality among cancer
survivors, allowing for analysis of overall and cause-specific death
patterns [@xu2022lasso]. The dataset includes variables such as age,
race, marital status, T stage, N stage, 6th stage, grade, A stage, tumor
size, estrogen status, progesterone status, regional nodes examined,
regional nodes positive, survival months, and status [@sembay2021seer].

Although LASSO regression has been valuable across many research areas,
it is particularly useful in medical studies where identifying the most
influential predictors is essential for improving patient outcomes. The
LASSO-Cox model is especially valuable for identifying predictors of
time-to-event outcomes in the presence of censored data. Using the SEER
dataset, this study applies LASSO-Cox regression—an L1-penalized Cox
proportional hazards model—for feature selection to identify key
clinical variables associated with breast cancer survival while reducing
overfitting and improving model stability [@lee2025machine].

## Methods

Survivial analysis involves the statistical modeling of the time until
an event, in this case death, occurs. Cox regression is widely used and
an effective method for performing survival analysis. The selected
demographic, clinical, and outcome variables were used as predictors in
a LASSO-Cox regression model to identify factors associated with patient
survival time.

Variable selection was performed using a Cox proportional hazards model
with an L1 (LASSO) penalty. In standard linear regression, LASSO is
defined as [@emmert2019high]:

$$ \hat{\beta} = arg \space \min \left\{\frac{1}{2n} \sum_{i=1}^{n} (y_i - \sum_{j} \beta_j x_{ij})^2 + \lambda||\beta||_1\right\}  $$

where $y_{i}$ represents the observed outcome, $x_{ij}$ denotes
predictor values, β is the vector of regression coefficients, and λ is a
tuning parameter which controls the strength of the penalty. The L1
penalty encourages sparsity by shrinking weaker coefficients exactly to
zero, thereby performing automatic variable selection.

For the LASSO-Cox model, coefficients are instead estimated by
maximizing the penalized Cox partial likelihood:

$$ \hat{\beta}_{\text{LASSO-Cox}} = \arg_{\beta}\max\left \{PL(\beta) - \lambda \sum_{j=1}^{p}|\beta_j| \right\} $$

where PL(β) denotes the Cox partial likelihood function and λ is a
positive tuning parameter that controls the amount of shrinkage
[@shahraki2015survival]. The L1 penalty shrinks some regression
coefficients to zero while estimating hazard ratios for the remaining
predictors.

**Assumptions:**

-   Hazard ratios between individuals remain constant over time
    (proportional hazards assumption)

-   Survival times of different patients are independent from one
    another

-   Censoring is assumed to be non-informative, meaning that the reason
    for censoring is unrelated to the probability of the event

The analysis was conducted in R (Version 4.4.2) using the glmnet package
which is designed for fitting penalized regression models, including
penalized Cox models. The strength of the penalty is controlled by λ. To
select an appropriate value of λ, k-fold cross-validation was used. This
procedure repeatedly partions the data into training and validation sets
to evaluate model performance and select the penalty that achieves an
optimal balance between model complexity and predictive accuracy.

Predictors with non-zero coefficients in the final LASSO-Cox model are
interpreted as significant variables associated with breast cancer
survival.

## Analysis and Results

### Data Exploration and Visualization

This study uses breast cancer data from the Surveillance, Epidemiology,
and End Results (SEER) program, which is maintained by the National
Cancer Institute (NCI). The dataset, composed of 4,024 patients,
includes demographic, clinical, and outcome variables for female
patients diagnosed with infiltrating ductal and lobular breast carcinoma
between 2006 and 2019. These variables are well suited for predictive
modeling of patient survival outcomes. The dataset excludes patients
with unknown tumor size, examined regional lymph nodes (LNs), regional
positive LNs, and patients whose survival months were less than 1 month
[@sembay2021seer].

| Category | Variables |
|------------------------------------|------------------------------------|
| Demographics | Age at diagnosis, race, marital status (married, single, widowed, divorced, separated) |
| Tumor Characteristics | T stage (T1–T4), N stage (N1–N3), 6th AJCC stage (IIA–IIIC), tumor grade (I–IV), A stage (regional/distant), tumor size |
| Hormone Receptors | Estrogen status, progesterone status (positive/negative) |
| Lymph Nodes | Nodes examined, nodes positive |
| Outcome | Survival months, vital status (alive/dead) |

: Table 1. Characteristics of the variables included in the study.

Using the variables defined above, analysis of the variables to predict
survival will be analyzed. To start, below is the distribution of age at
the time of breast cancer diagnosis.

```{r}
#install.packages("survminer")
#install.packages("glmnet")
library(glmnet)
library(survminer)
library(survival)
library(ggplot2)
library(corrplot)
seer_data <- read.csv("SEER_Breast_Cancer_Data.csv")
```

```{r}
ggplot(seer_data, aes(x = Age)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "black", alpha = 0.7) +
  labs(title = "Age at Diagnosis",
       x = "Age (Years)",
       y = "Frequency") +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

```

Graph #1. Distribution of age at breast cancer diagnosis.

This histogram shows that patients are between ages 30 to 70 with a
majority of diagnoses aged 50 and older. Due to the data that has been
excluded in this study, the age of the women that have been diagnosed
with breast cancer could vary if it were to be included. The next
analysis shows a boxplot of survival based on stages of breast cancer.

```{r}
ggplot(seer_data, aes(x = X6th.Stage, y = Survival.Months)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Survival by AJCC Stage",
       x = "6th AJCC Stage",
       y = "Survival Months") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) 

```

Graph #2. Distribution of survival time across AJCC stages.

The boxplot shows that there is a general downward trend in survival
months based on 6th AJCC cancer stages (IIA, IIB, IIIA, IIIB, IIIC). The
early stages of breast cancer (IIA, IIB, IIIA) show a higher median
survival where later stages (especially IIIC) show a lower median
survival. This is suggesting that a higher stage breast cancer diagnosis
is associated with shorter survival. Although, the median survival
across the stages do not differ dramatically which shows that there is
overlap between groups, thus stage alone may not fully explain a
survival outcome. Due to the length of the whiskers, the length of
survival varies widely within each stage and there may be other
predictors that likely influence survival.

```{r}
seer_data$event <- ifelse(seer_data$Status == "Dead", 1, 0)

fit <- survfit(Surv(Survival.Months, event) ~ 1, data = seer_data)

plot <- ggsurvplot(fit,
                   data = seer_data,
                   conf.int = TRUE,
                   censor = FALSE,
                   risk.table = TRUE,
                   size = 1.2, 
                   pval = FALSE,
                   xlab = "Time (Months)",
                   ylab = "Survival Probability",
                   title = "Kaplan–Meier Survival Curve",
                   palette = "blue")

plot$plot <- plot$plot +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

plot


```

Graph #3. Kaplan-Meier estimate of overall survival among breast cancer
patients.

A Kaplan–Meier survival analysis was conducted to estimate the overall
survival over a 100-month follow-up period. The survival curve indicates
a gradual decline in survival probability over time, beginning at 100%
at baseline and decreasing steadily to approximately 78% by 100 months.
The number at risk decreased over time, from 4,024 participants at
baseline to 444 at 100 months, reflecting both events and censoring.
Overall, the results suggest relatively high survival across the study
population with a slow and consistent reduction in survival probability
over time.

```{r}
seer_data$event <- ifelse(seer_data$Status == "Dead", 1, 0)
seer_data$X6th.Stage <- as.factor(seer_data$X6th.Stage)

fit <- survfit(Surv(Survival.Months, event) ~ X6th.Stage, data = seer_data)

plot <- ggsurvplot(fit, 
                   data = seer_data, 
                   conf.int = TRUE, 
                   risk.table = TRUE, 
                   risk.table.height = .40, 
                   risk.table.fontsize = 3.5, 
                   pval = FALSE, 
                   xlab = "Time (Months)",
                   ylab = "Survival Probability",
                   title = "Kaplan–Meier Survival Curves by 6th Stage",
                   legend.labs = c("IIA","IIB","IIIA","IIIB","IIIC"))
plot$plot <- plot$plot + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

plot

```

Graph #4 Kaplan–Meier Survival by Hormone Receptor Status

This graph shows Kaplan–Meier survival curves grouped by the 6th stage
classification. The curves illustrate how survival probability changes
over time for patients diagnosed at different cancer stages. Patients
diagnosed at earlier stages (IIA and IIB) generally show higher survival
probabilities throughout the follow-up period. In contrast, patients
diagnosed at later stages (IIIA, IIIB, and IIIC) experience a faster
decline in survival probability over time.

Among the later stages, stage IIIC appears to have the lowest survival
probability, while stage IIIA and IIIB fall between the early and
advanced stages. Overall, the curves suggest that patients diagnosed
with more advanced stages tend to have lower survival probabilities
compared to those diagnosed at earlier stages.

The risk table below the graph shows the number of patients remaining at
risk at different time points during the study period.

```{r}
numeric_data <- seer_data[, c("Age","Tumor.Size","Regional.Node.Examined","Reginol.Node.Positive","Survival.Months")]
numeric_data <- data.frame(lapply(numeric_data, as.numeric))
cor_matrix <- cor(numeric_data, use = "complete.obs")
#print(cor_matrix)
corrplot(cor_matrix, method = "color", type = "upper", addCoef.col = "black",tl.col = "black",tl.srt = 45,number.cex = 0.7,
        title = "Correlation Between Continuous Variables", mar = c(0,0,2,0))

```

Graph #5: Correlation between continuous variables.

This heatmap illustrates the correlations between age, tumor size,
regional nodes examined, regional nodes positive, and survival months.
The strongest relationship is observed between the number of regional
lymph nodes examined and the number of positive lymph nodes. With a
correlation coefficient of r = 0.41, this indicates that patients with
more lymph nodes examined tend to have more positive nodes detected.
Tumor size and the number of positive lymph nodes show a small to
moderate positive correlation (r = 0.24), suggesting that larger tumors
may be associated with increased lymph node involvement. A weak negative
correlation is observed between positive lymph nodes and survival months
(r = -0.14), indicating that greater lymph node involvement may be
associated with shorter survival time. There appears to be little to no
relationship between age and survival months (r = -0.01) in this
dataset.

```{r}
ggplot(seer_data, aes(x = Status, y = Tumor.Size, fill = Status)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3) +
  labs(
    title = "Distribution of Tumor Size by Survival Status",
    x = "Survival Status",
    y = "Tumor Size"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5)
  )
```

Graph #6: Distribution of Tumor Size by Survival Status

The boxplot shows that there is a correlation between tumor size and
survival status, as patients with larger tumor sizes tend to die
compared to those with smaller tumor sizes. The Alive group shows a
lower median tumor size, whereas the Dead group shows a higher median
tumor size. There is overlap between the two groups, indicating that
tumor size alone is not a strong predictor of survival in cancer
patients. The wider spread and presence of several large outliers,
particularly among deceased patients, suggest considerable variability
and the influence of additional prognostic factors.

## Modeling and Results

To identify key predictors of breast cancer survival among patients in
the SEER dataset, a LASSO-penalized Cox proportional hazards model was
used. The data were first prepared for use in the LASSO-Cox model.

Categorical predictors were converted into indicator (dummy) variables
so they could be included in the model. Continuous variables were
standardized automatically within the glmnet framework to ensure that no
single variable dominated the L1 penalty applied by LASSO. Because the
outcome of interest was time to event, a penalized Cox proportional
hazards model was used. Observations with missing values were removed to
maintain consistency in the analysis.

```{r}
#remove observations with missing values
vars <- c("Age", "Race", "Marital.Status", "T.Stage", "N.Stage", "X6th.Stage", "Grade", "A.Stage", "Tumor.Size", "Estrogen.Status", "Progesterone.Status", "Regional.Node.Examined", "Reginol.Node.Positive", "Survival.Months", "Status")

#Subset the dataset to include only the variables used in the survival analysis

model_data <- seer_data[, vars]

#Remove observations with missing values
model_data <- na.omit(model_data)

#Create an event indicator variable required for survival analysis
# 1 indicates the event occurred (death) and 0 indicates censored observations (alive)
model_data$event <- ifelse(model_data$Status == "Dead", 1, 0)

```

A survival object was created using the survival time and event
indicator. Predictor variables were converted into a model matrix to
provide a numeric input suitable for the glmnet package, with
categorical variables represented through dummy coding. A penalized Cox
proportional hazards model (LASSO) was then fit using glmnet which
estimated regression coefficients while applying regularization to
perform variable selection.

```{r}
y <- Surv(model_data$Survival.Months, model_data$event)

#x <- model.matrix(Surv(Survival.Months, event) ~ Age + Race + Marital.Status + T.Stage + N.Stage + X6th.Stage + Grade + A.Stage + Tumor.Size + Estrogen.Status + Progesterone.Status + Regional.Node.Examined + Reginol.Node.Positive, data = model_data)[,-1]

x <- model.matrix(event ~ . - event - Status - Survival.Months, data = model_data)[,-1]


#Fit a penalized cox regression model using glmnet
fit <- glmnet(x, y, family = "cox")
plot(fit)
```

The above coefficient path plot demonstrates how predictor effects
evolve as the penalty parameter (λ) decreases. The x-axis, shown as
-log(λ), represents the strength of the penalty, with the left side
corresponding to stronger regularizations and a simpler model, and the
right side corresponding to weaker regularization and a more complex
model with additional variables. The y-axis displays the estimated
coefficients, indicating the magnitude and direction of each predictor's
effects. A negative value represents a decreased hazard where a positive
value annotates an increased hazard. The numbers along the top of the
model represent the number of non-zero coefficients at each value of λ,
demonstrating that more variables enter the model as the penalty
decreases.

The function 'cv.glmnet()' was used to determine the optimal value of
the penalty parameter (λ) through k-fold cross-validation. This approach
evaluates model performance across a sequence of λ values by estimating
the cross-validation partial liklihood deviance. The resulting plot
displays the mean cross-validation error and associated confidence
intervals for each λ.

```{r}
#Perform cross-validation to identify the optimal lambda value
cvfit <- cv.glmnet(x, y, family = "cox", alpha = 1)

#Plot the cross-validation error across lambda values
plot(cvfit)
```

The optimal value of λ identified through cross-validation is determined
by cvfit\$lambda.min. The coefficients corresponding to the optimal λ
value are then extracted to identify predictors with non-zero
coefficients that remain in the final LASSO-Cox model.

```{r}
cvfit$lambda.min
coef(cvfit, s = "lambda.min")
```

The following coefficients represent predictors with non-zero values
selected by the LASSO-Cox proportional hazards model and are interpreted
as key factors influencing survival outcomes.

```{r}
lasso_coef <- coef(cvfit, s = "lambda.min")
lasso_coef[lasso_coef != 0]
```

The final LASSO-penalized Cox model identified 22 predictors with
non-zero coefficients at the optimal penalty parameter, highlighting the
most influential variables associated with survival outcomes. The
variables with positive coefficients were linked to an increased hazard,
or worse survival, while negative coefficients were linked to a
decreased hazard, or better survival.

## Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

ChatGPT was used to assist with grammar and vocabulary corrections.

## References
