---
title: "LASSO Regression - Spring 2026 "
subtitle: "Determining Survival Rates for Breast Cancer Patients"
author: "Stephanie Hopkins (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)

## Introduction

The LASSO, Least Absolute Shrinkage and Selector Operator, regression
model has become very popular over the last 30 years. The model was
first introduced by Robert Tibshirani to perform variable selection and
regularization to improve prediction accuracy and interpretability
[@emmert2019high]. Since its introduction, LASSO has been used to
research topics in a numerous amount of fields including biology,
computational social sciences, social media, the stock market, economy,
political science, robotics, climatology, and pharmacology
[@emmert2019high]. In LASSO regression, the prediction error is
minimized by identifying the variables and corresponding regression
coefficients and then putting constraints on the model which pushes the
regression coefficients to zero [@ranstam2018lasso]. Due to this, this
regression model works extremely well with outliers, thus dealing with
large-scale problems very efficiently [@roth2004generalized].

The goal of this analysis is to use LASSO regression to identify factors
that are most strongly associated with survival in breast cancer
patients. "Breast cancer is the most frequent cancer in women and is the
second most common cancer across the globe" [@usman2021comparing].
Understanding which clinical and demographic factors are most predictive
in patient survival may help to improve prognostic modeling and support
more informed clinical decision-making. Using data from the SEER
dataset, this analysis will use a Cox proportional hazards model to
perform variable selection and identify key predictors of breast cancer
survival.

## Methods

-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.

LASSO regression can be defined by [@emmert2019high]:

$$
\hat{\beta} = arg \space \min \left\{\frac{1}{2n} \sum_{i=1}^{n} (y_i - \sum_{j} \beta_j x_{ij})^2 + \lambda||\beta||_1\right\} 
$$

where (define variables)

## Analysis and Results

### Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

A study was conducted to determine how...

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{python}
import pandas as pd
```

```{r, warning=FALSE, echo=TRUE}
# Load Data
kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

### Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
