---
title: "LASSO Regression - Spring 2026 "
subtitle: "Determining Predictors in Survival for Breast Cancer Patients"
author: "Stephanie Hopkins & Aminata Gado Oumarou (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)

## Introduction

The LASSO (Least Absolute Shrinkage and Selection Operator) regression
model has become increasingly popular over the past 30 years. It was
first introduced by Robert Tibshirani as a method for variable selection
and regularization to improve prediction accuracy and interpretability.
Since then, LASSO has been widely applied across many fields, including
biology, computational social sciences, social media, the stock market,
economics, political science, robotics, climatology, and pharmacology
[@emmert2019high].

Regression is a fundamental task in supervised machine learning, used to
model relationships between variables and address key challenges such as
parameter estimation and variable selection [@muthukrishnan2016lasso].
As machine learning methods continue to grow in popularity and show
strong results, their importance in research has increased. In medicine
especially, predictive models developed through machine learning can
help professionals distinguish between risk groups, design personalized
treatment plans, and improve patient survival outcomes
[@huang2023practical].

LASSO regression works by minimizing prediction error while applying a
constraint that shrinks some regression coefficients toward zero
[@ranstam2018lasso]. It does this by adding an L1 penalty term to the
loss function, which constrains the sum of the absolute values of the
coefficients [@ranstam2018lasso]. This penalty forces weaker predictors
to exactly zero, effectively performing automatic variable selection
[@usman2021comparing]. While traditional LASSO in linear regression
minimizes the residual sum of squares subject to an L1 constraint
[@tibshirani1996regression], the method can be extended to survival
analysis through the Cox proportional hazards model. In survival
analysis, the Cox proportional hazards model is widely used to examine
the relationship between covariates and time-to-event outcomes, such as
overall survival. Rather than minimizing residual error, the Cox model
estimates regression coefficients by maximizing the partial likelihood
function. The LASSO-Cox model incorporates an L1 penalty into this
partial likelihood, allowing some regression coefficients to shrink
exactly to zero while estimating hazard ratios [@tibshirani1997lasso].
This approach reduces multicollinearity, improves estimation stability,
and produces a sparse and interpretable model, which is especially
valuable in high-dimensional clinical datasets [@shahraki2015survival].

The goal of this analysis is to use LASSO-Cox regression to identify the
factors most strongly associated with breast cancer survival time.
Cancer continues to have a major impact worldwide. In 2025, an estimated
2,041,910 new cancer cases will be diagnosed in the United States, and
approximately 618,120 people are expected to die from the disease
[@nci2025statistics]. Breast cancer remains a significant public health
concern. By 2019, more than 3.8 million individuals in the United States
were living with breast cancer, and it is still the most commonly
diagnosed cancer among women and the second most common cancer globally
[@usman2021comparing]. Breast cancer begins in the breast’s tissues and
develops in the epithelial cells of the ducts (85% of cases) and lobules
(15%). Causes include hereditary factors, age, postmenopausal obesity,
alcohol or smoking, and other factors [@hadba2024comparison].
Identifying the clinical and demographic factors most strongly linked to
survival can help improve prognostic models and support better clinical
decision-making. Previous research has identified numerous factors
associated with breast cancer prognosis and survival. These factors are
commonly grouped into three categories: (1) demographic and genetic
characteristics, such as age at diagnosis, marital status, and
reproductive history; (2) clinicopathological factors, including tumor
location, size, and histologic grade; and (3) treatment-related
variables, such as drug therapy, radiotherapy, and chemotherapy
[@teng2022dynamic].

To accomplish this, the analysis uses the Surveillance, Epidemiology,
and End Results (SEER) dataset to determine the most important variables
associated with breast cancer survivability. SEER has collected and
maintained high-quality, validated data on mortality among cancer
survivors, allowing for analysis of overall and cause-specific death
patterns [@xu2022lasso]. The dataset includes variables such as age,
race, marital status, T stage, N stage, 6th stage, grade, A stage, tumor
size, estrogen status, progesterone status, regional nodes examined,
regional nodes positive, survival months, and status [@sembay2021seer].

Although LASSO regression has been valuable across many research areas,
it is particularly useful in medical studies where identifying the most
influential predictors is essential for improving patient outcomes. The
LASSO-Cox model is especially valuable for identifying predictors of
time-to-event outcomes in the presence of censored data. Using the SEER
dataset, this study applies LASSO-Cox regression—an L1-penalized Cox
proportional hazards model—for feature selection to identify key
clinical variables associated with breast cancer survival while reducing
overfitting and improving model stability [@lee2025machine].

## Methods

The selected demographic, clinical, and outcome variables were used as
predictors in a LASSO-Cox regression model to identify factors
associated with patient survival time.

The data were first prepared for use in the LASSO-Cox model. Categorical
predictors were converted into indicator (dummy) variables so they could
be included in the model. Continuous variables were scaled to common
ranges to ensure that no single variable dominated the L1 penalty
applied by LASSO. Because the outcome of interest was time to event, a
penalized Cox proportional hazards model was used. Observations with
missing values were removed to maintain consistency in the analysis.

Variable selection was performed using a Cox proportional hazards model
with an L1 (LASSO) penalty. In standard linear regression, LASSO is
defined as [@emmert2019high]:

$$ \hat{\beta} = arg \space \min \left\{\frac{1}{2n} \sum_{i=1}^{n} (y_i - \sum_{j} \beta_j x_{ij})^2 + \lambda||\beta||_1\right\}  $$

where $y_{i}$ represents the observed outcome, $x_{ij}$ denotes
predictor values, β is the vector of regression coefficients, and λ is a
tuning parameter which controls the strength of the penalty. The L1
penalty encourages sparsity by shrinking weaker coefficients to zero,
thereby performing automatic variable selection.

For the LASSO-Cox model, coefficients are instead estimated by
maximizing the penalized Cox partioal likelihood:

$$ \hat{\beta}_{\text{LASSO-Cox}} = \arg_{\beta}\max\left \{PL(\beta) - \lambda \sum_{j=1}^{p}|\beta_j| \right\} $$

where PL(β) denotes the Cox partial likelihood function and λ is a
positive tuning parameter that controls the amount of shrinkage
[@shahraki2015survival]. The L1 penalty shrinks some regression
coefficients to zero while estimating hazard ratios for the remaining
predictors.

The analysis was conducted in R using the glmnet package which is
designed for fitting penalized regression models, including penalized
Cox models. The strength of the penalty is controlled by λ. To select an
appropriate value of λ, k-fold cross-validation sets to evaluate model
performance and select the penalty that achieves an optimal balance
between model complexity and predictive accuracy.

Predictors with non-zero coefficients in the final LASSO model were
interpreted as significant variables associated with breast cancer
survival.

## Analysis and Results

### Data Exploration and Visualization

This study uses breast cancer data from the Surveillance, Epidemiology,
and End Results (SEER) program, which is maintained by the National
Cancer Institute (NCI). The dataset, composed of 4,024 patients,
includes demographic, clinical, and outcome variables for female
patients diagnosed with infiltrating ductal and lobular breast carcinoma
between 2006 and 2019. These variables are well suited for predictive
modeling of patient survival outcomes. The dataset excludes patients
with unknown tumor size, examined regional lymph nodes (LNs), regional
positive LNs, and patients whose survival months were less than 1 month
[@sembay2021seer].

| Category | Variables |
|------------------------------------|------------------------------------|
| Demographics | Age at diagnosis, race, marital status (married, single, widowed, divorced, separated) |
| Tumor Characteristics | T stage (T1–T4), N stage (N1–N3), 6th AJCC stage (IIA–IIIC), tumor grade (I–IV), A stage (regional/distant), tumor size |
| Hormone Receptors | Estrogen status, progesterone status (positive/negative) |
| Lymph Nodes | Nodes examined, nodes positive |
| Outcome | Survival months, vital status (alive/dead) |

: Table 1. Characteristics of the variables included in the study.

Using the variables defined above, analysis of the variables to predict
survival will be analyzed. To start, below is the distribution of age at
the time of breast cancer diagnosis.

```{r}
library(ggplot2)
data <- read.csv("SEER_Breast_Cancer_Data.csv")
```

```{r}
ggplot(data, aes(x = Age)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "black", alpha = 0.7) +
  labs(title = "Age at Diagnosis",
       x = "Age (Years)",
       y = "Frequency") +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

```

This histogram shows that patients are between ages 30 to 70 with a
majority of diagnoses aged 50 and older. Due to the data that has been
excluded in this study, the age of the women that have been diagnosed
with breast cancer could vary if it were to be included. The next
analysis shows a boxplot of survival based on stages of breast cancer.

```{r}
ggplot(data, aes(x = X6th.Stage, y = Survival.Months)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Survival by AJCC Stage",
       x = "6th AJCC Stage",
       y = "Survival Months") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) 

```

The boxplot shows that there is a general downward trend in survival
months based on 6th AJCC cancer stages (IIA, IIB, IIIA, IIIB, IIIC). The
early stages of breast cancer (IIA, IIB, IIIA) show a higher median
survival where later stages (especially IIIC) show a lower median
survival. This is suggesting that a higher stage breast cancer diagnosis
is associated with shorter survival. Although, the median survival
across the stages do not differ dramatically which shows that there is
overlap between groups, thus stage alone may not fully explain a
survival outcome. Due to the length of the whiskers, the length of
survival varies widely within each stage and there may be other
predictors that likely influence survival.

```{r}
#install.packages("survminer")
library(survminer)
library(survival)

data$event <- ifelse(data$Status == "Dead", 1, 0)

fit <- survfit(Surv(Survival.Months, event) ~ 1, data = data)

plot <- ggsurvplot(fit,
                   data = data,
                   conf.int = TRUE,
                   risk.table = TRUE,
                   pval = FALSE,
                   xlab = "Time (Months)",
                   ylab = "Survival Probability",
                   title = "Kaplan–Meier Survival Curve",
                   palette = "blue")

plot$plot <- plot$plot +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

plot


```

A Kaplan–Meier survival analysis was conducted to estimate the overall
survival over a 100-month follow-up period. The survival curve indicates
a gradual decline in survival probability over time, beginning at 100%
at baseline and decreasing steadily to approximately 80% by 100 months.
The stepwise pattern reflects the occurrence of events throughout the
study period, while censoring is indicated by tick marks along the
curve. The number at risk decreased over time, from 4,024 participants
at baseline to 444 at 100 months, reflecting both events and censoring.
Overall, the results suggest relatively high survival across the study
population with a slow and consistent reduction in survival probability
over time.

```{r}
ggplot(data, aes(x = Status, y = Tumor.Size, fill = Status)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3) +
  labs(
    title = "Distribution of Tumor Size by Survival Status",
    x = "Survival Status",
    y = "Tumor Size"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5)
  )
```

The boxplot shows that there is a correlation between tumor size and
survival status, as patients with larger tumor sizes tend to die
compared to those with smaller tumor sizes. The Alive group shows a
lower median tumor size, whereas the Dead group shows a higher median
tumor size. There is overlap between the two groups, indicating that
tumor size alone is not a strong predictor of survival in cancer
patients. The wider spread and presence of several large outliers,
particularly among deceased patients, suggest considerable variability
and the influence of additional prognostic factors.

 

## Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

## Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
