---
title: "LASSO Regression - Spring 2026 "
subtitle: "Determining Predictors in Survival for Breast Cancer Patients"
author: "Stephanie Hopkins & Aminata Gado Oumarou (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)

## Introduction

The LASSO, Least Absolute Shrinkage and Selector Operator, regression
model has become very popular over the last 30 years. The model was
first introduced by Robert Tibshirani to perform variable selection and
regularization to improve prediction accuracy and interpretability .
Since its introduction, LASSO has been used to research topics in
numerous fields including biology, computational social sciences, social
media, the stock market, economy, political science, robotics,
climatology, and pharmacology [@emmert2019high]. In LASSO regression,
the prediction error is minimized by identifying the variables and
corresponding regression coefficients and then putting constraints on
the model which pushes the regression coefficients to zero
[@ranstam2018lasso]. Those coefficients are then eliminated from the
model reducing its complexity.

LASSO reduces model complexity by adding an L1 penalty term added to the
loss function which constrains the sum of the absolute values of the
regression coefficients [@ranstam2018lasso]. This penalty is what
shrinks the weaker predictors to zero and is effectively performing
automatic variable selection [@usman2021comparing]. Due to this, the
LASSO regression model can still work with outliers, thus dealing with
large-scale problems very efficiently [@roth2004generalized]. By
reducing overfitting and emphasizing the most relevant predictors, LASSO
produces a more interpretable model which is key in clinical survival
modeling where interpretability and reproducibility are critical for
predicting patient outcomes [@ranstam2018lasso].

The goal of this analysis is to use LASSO regression to identify factors
most strongly associated with survival in breast cancer patients. Cancer
has a substantial impact on society worldwide. In 2025, an estimated
2,041,910 new cancer cases will be diagnosed in the United States, and
approximately 618,120 people are expected to die from the disease
[@siegel2025cancer].

Breast cancer represents a particularly significant public health
concern. By 2019, the number of individuals living with breast cancer in
the United States was estimated to exceed 3.8 million, and it remains
the most frequently diagnosed cancer among women and the second most
common cancer globally [@usman2021comparing]. Understanding which
clinical and demographic factors are most predictive in patient survival
may help to improve prognostic modeling and support more informed
clinical decision-making. Using the Surveillance, Epidemiology, and End
Results (SEER) dataset, this analysis will determine the most prominent
variables in breast cancer survivability. This dataset includes the
variables age, race, marital status, T stage, N stage, 6th stage, grade,
A stage, tumor size, estrogen status, progesterone status, regional node
examined, regional node positive, survival months, and status.

These variables will now be further defined from the SEER dataset
[@Teng2019SEER] . Age is the age at diagnosis, race is the race of the
patient, and marital status at the time of diagnosis includes married
(including common law), single (never married), widowed, divorced and
separated. T stage is the size of the tumor where T1 is tumor ≤ 2 cm, T2
is tumor \> 2 cm but ≤ 5 cm, T3 is tumor \> 5cm, and T4 is a tumor that
is any size which has directly invaded the chest wall, skin, or is
classified as inflammatory. N stage is the number of lymph nodes
involved where N1 is 1-3 positive lymph nodes, N2 is 4-9 positive lymph
nodes, and N3 is 10+ positive nodes or specific node groups. 6th stage
is the overall stage of cancer which includes IIA, IIB, IIC, IIIA, IIIB,
and IIIC. Next is the grade of the breast cancer which is broken down
into grade I, grade II, grade III, and grade IV. A stage is defined by
either regional or distant. Tumor size is the size of the tumor at the
time of diagnosis. Estrogen status is the status of the estrogen
receptor which is defined as either negative or positive. Progesterone
status is the status of the progesterone receptor which is again either
negative or positive. Regional node examined is whether the regional
lymph nodes were examined and if so then how many. Regional node
positive states how many of the regional lymph nodes were positive for
cancer. Survival months states the number of months the patient survived
after diagnosis. Lastly, status states whether the patient is alive or
passed away.

While LASSO regression has proven valuable across many scientific
research areas, it is particularly well suited for medical research
where identifying the most influential predictors is essential for
improving patient outcomes. Using this data, the analysis applies LASSO
regression to perform variable selection on the defined variables and
identify the key predictors of breast cancer survival.

ChatGPT was used to help review grammar and clarity of this
introduction.

## **Article Reviews:**

**High-Dimensional LASSO-Based Computational Regression Models**

This article summarizes high-dimensional LASSO-based computational
regression models as an effective solution for situations in which the
number of predictors is large relative to the sample size. The defining
feature of LASSO is the use of an L1 penalty that performs coefficient
shrinkage and variable selection which reduces weak or irrelevant
predictors to a coefficient of zero. This regression model is good for
biomedical, genomic, engineering applications where multicollinearity
and overfitting are common challenges. It highlights how LASSO improves
model interpretability, numerical stability, and predictive performance
compared to unregularized regression by controlling variance without
sacrificing key signal variables. Overall, it is shown that LASSO is
used for high-dimensional regression, offering a balance between model
complexity and predictive accuracy.

**LASSO Regression**

LASSO regression is a great method for addressing issues that other
models tend to struggle with, like overfitting or being overly
optimistic about their predictions. The model minimizes prediction error
by putting constraints on the model parameters which forces coefficients
towards zero. Then, the coefficients that are zero are removed from the
model. This limit is controlled by a value called λ. A benefit of doing
this is that it reduces overfitting in the model. A disadvantage of the
model is that because the focus in on the best combined prediction, it
may not be reliable in terms of independent risk factors. Although LASSO
regression is a strong model that often outperforms other methods, it is
not perfect and may miss some subtle patterns in the data.

**The Generalized LASSO**

This article summarizes what the least absolute shrinkage and selection
operator (LASSO) regression model is. It emphasizes LASSO's role as a
regularization method that automaticlaly performs variable selection by
shrinking many coefficients exactly to zero, reducing model complexity
without sacrificing predictive accuracy. It can be optimized to
guarantee convergence which makes it good to use for large-scale
regression problems. Generalized LASSO maintains sparsity while
improving computational scalability and robustness to outliers. Its
effectiveness is highlighted by performing well with fewer active
variables which is highly effective in high-dimensional regression.

**Comparing the Prediction Accuracy of Ridge, LASSO and Elastic Net
Regression Models with Linear Regression Using Breast Cancer Data**

This paper used 3 different methods to determine factors that can help
determine the survivability of breast cancer patients. This is important
because it can help determine what factors can predict the survivability
of breast cancer. LASSO had the lowest prediction error and most
effective variable selection. Of the 14 variables examined, 8 of them
were significant in predicting breast cancer. The LASSO model identified
seven key predictors of survival, making it the most precise. Patients
in the age range of 60-90, that are in stage 2 and stage 3 of the
disease have a lower survival time. One of the limitations in this paper
is that it had a smaller number of patients that it used in
Zaria-Nigeria so it was recommended to look at a broader range of
data.  

**Identification of genes associated with survival of breast cancer\
patients**

This study looked at different microRNA and mRNA data to predict
survivability in breast cancer. LASSO regression was used to identify 10
out of 56 microRNA's with non-zero coefficients that are associated with
prognosis of breast cancer patients. This is important because it can
help improve risk stratification, prognosis, and personalized treatment
planning for breast cancer patients. LASSO regression was the best
method for working with this data due to it being high-dimensional and
noisy. It was solved by classifying patients into high and low risk
survival groups. This research also included a single limited public
dataset, so additional biological experiments should be done. 

**Development of a prediction model based on LASSO regression to
evaluate the risk of non‑sentinel lymph node metastasis in Chinese
breast cancer patients with 1–2 positive sentinel lymph nodes**

This paper is similar to what we are researching although it's strictly
looking at lymph nodes as a factor in breast cancer patients. This paper
is important because it is specifically looking at Chinese patients, so
it covers Western populations and how care may be limited for Eastern
patients. LASSO regression was applied to select the most informative
predictors and build the risk score model. The model performance was
evaluated by looking at the area under the receiver operating
characteristic curve (AUC). LASSO performed well and found that 30% of
patients could possibly avoid axillary lymph node dissection
(ALND). This study was conducted at one research center so it is
recommended that this study be applied to a larger group to confirm
generalization. Overall, this approach can help doctors decide who
actually needs more invasive lymph node surgery and who can safely avoid
it. 

**Comparison of lasso logistic regression, artificial neural networks,
and support vector machine in predicting breast cancer**

This study evaluates LASSO Logistic Regression as a tool for classifying
breast tumors as benign or malignant and compares its performance
against other regression models using the Wisconsin Diagnostic Breast
Cancer (WDBC) dataset. LASSO was used by adding the L1 penalty to the
function to find the strongest variables. This is shown to work well in
high-dimensional medical datasets where the data can be highly
correlated. The data was split into 80% training and 20% testing sets.
LASSO outpreformed all of the other logistic regression models that it
was compared to.

**A Machine Learning Model for Predicting Breast Cancer Recurrence and
Supporting Personalized Treatment Decisions Through Comprehensive
Feature Selection and Explainable Ensemble Learning**

This study was conducted using the LASSO regression model to predict
breast cancer recurrence and support personalized treatment decisions.
Data were collected from 1,131 breast cancer patients using the health
record system at Kaohsiung Medical University Hospital. Using key
clinical and predictor factors, the model identified patients at higher
risk of recurrence. The study also used SHAP analysis to explain the
model results and support clinical decision-making. This model
outperformed a single model previously used for the same purpose.

**Logistic LASSO Regression for Dietary Intakes and Breast Cancer**

The goal of this study was to use LASSO regression to examine the
relationship between dietary intakes and breast cancer in women aged 50
and older. The model looked at 29 dietary variables alongside
well-established breast cancer risk factors. This research is important
because it could help women, especially those at higher risk, such as
BRCA gene carriers, identify foods to avoid. The study used NHANES
survey data and R software. LASSO regression with ten-fold
cross-validation was employed to select important predictors while
avoiding overfitting. Overall, there was no strong correlation between
dietary intake or restrictions and time since diagnosis, except for
alcohol, which showed a positive association. A statistically
significant association was found between higher vitamin B12 intake and
breast cancer.

**A LASSO-based survival prediction model for patients with synchronous
colorectal carcinomas based on SEER**

This study uses LASSO-based Cox regression to model and predict the
overall survival of patients with synchronous colorectal carcinomas. The
data came from the SEER database, which included patients diagnosed with
colorectal cancer more than once between 2004 and 2013. The LASSO-Cox
model was compared to the standard Cox model using the same dataset to
determine if the LASSO-based regression produced a more accurate
prediction. The LASSO-based Cox model performed better by selecting the
most important variables while reducing overfitting. It showed that
certain patient and tumor characteristics, such as age, tumor depth,
sex, and whether chemotherapy was given, are important predictors of
survival. The LASSO model predicted that older men with T3 tumors are at
higher risk, and the model’s decision-curve analysis suggests it can
guide clinical decisions to improve patient outcomes.

**A practical recurrence risk model based on Lasso-Cox regression for
gastric cancer**

A dataset of 1,851 patients with gastric cancer was used to build models
that predict recurrence after treatment and help improve patient
outcomes. Several methods including LASSO regression, Cox regression,
and nomogram analysis were used for selecting variables, constructing
the models, and validating their performance so the results could be
compared. LASSO regression helped reduce overfitting and
multicollinearity by keeping only the most relevant factors, while the
other approaches included more variables. The findings showed that
pathological stage, tumor size, lymph node status, intraoperative blood
loss, and AFP and CA199 levels were strongly linked to gastric cancer
recurrence after radical gastrectomy, and the final model showed good
accuracy and calibration. The multivariable Cox model also performed
well and could be useful for guiding more personalized treatment plans.

**Penalized Cox regression analysis in the high-dimensional and
low-sample size settings, with applications to microarray gene
expression data**

The main goal of this paper is to adapt the LASSO penalization to the
Cox proportional hazards framework since this type of analysis works
best for predicting survivability.  It is shown to be a more reliable
model for survival outcomes when the number of predictors is large to
the number of observations.  The paper addresses how to build
interpretable survival models when the data are high-dimensional and
noisy.  This is shown to be relevant in modern cancer research. The
study extends the LASSO penalization to the Cox proportional hazards
framework by adding an L1 penalty to the Cox partial likelihood. It was
shown that it has a strong predictive performance and effective variable
selection.  Limitations of this study was that performance can degrade
when predictors are highly correlated and that LASSO can mis predictors
with small but real effects. 

## Methods

-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.

LASSO regression can be defined by [@emmert2019high]:

$$
\hat{\beta} = arg \space \min \left\{\frac{1}{2n} \sum_{i=1}^{n} (y_i - \sum_{j} \beta_j x_{ij})^2 + \lambda||\beta||_1\right\} 
$$

where (define variables)

## Analysis and Results

### Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

A study was conducted to determine how...

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{python}
import pandas as pd
```

```{r, warning=FALSE, echo=TRUE}
# Load Data
kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

### Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
