---
title: "LASSO Regression - Spring 2026 "
subtitle: "Determining Predictors in Survival for Breast Cancer Patients"
author: "Stephanie Hopkins (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)

## Introduction

The LASSO, Least Absolute Shrinkage and Selector Operator, regression
model has become very popular over the last 30 years. The model was
first introduced by Robert Tibshirani to perform variable selection and
regularization to improve prediction accuracy and interpretability
[@emmert2019high]. Since its introduction, LASSO has been used to
research topics in numerous fields including biology, computational
social sciences, social media, the stock market, economy, political
science, robotics, climatology, and pharmacology [@emmert2019high]. In
LASSO regression, the prediction error is minimized by identifying the
variables and corresponding regression coefficients and then putting
constraints on the model which pushes the regression coefficients to
zero [@ranstam2018lasso]. Those coefficients are then eliminated from
the model reducing its complexity.

LASSO reduces model complexity by adding an L1 penalty term added to the
loss function which constrains the sum of the absolute values of the
regression coefficients [@ranstam2018lasso]. This penalty is what
shrinks the weaker predictors to zero and is effectively performing
automatic variable selection [@usman2021comparing]. Due to this, LASSO
regression model can still work with outliers, thus dealing with
large-scale problems very efficiently [@roth2004generalized]. By
reducing overfitting and emphasizing the most relevant predictors, LASSO
produces a more interpretable model which is key in clinical survival
modeling where interpretability and reproducibility are critical for
predicting patient outcomes [@ranstam2018lasso].

The goal of this analysis is to use LASSO regression to identify factors
that are most strongly associated with survival in breast cancer
patients. "Breast cancer is the most frequent cancer in women and is the
second most common cancer across the globe" [@usman2021comparing].
Understanding which clinical and demographic factors are most predictive
in patient survival may help to improve prognostic modeling and support
more informed clinical decision-making. Using the Surveillance,
Epidemiology, and End Results (SEER) dataset, this analysis will
determine the most prominent variables in breast cancer survivability.
This dataset includes the variables age, race, marital status, T stage,
N stage, 6th stage, grade, A stage, tumor size, estrogen status,
progesterone status, regional node examined, regional node positive,
survival months, and status.

These variables will now be further defined. Age is the age at
diagnosis, race is the race of the patient, and marital status at the
time of diagnosis includes married (including common law), single (never
married), widowed, divorced and separated. T stage is the size of the
tumor where T1 is tumor ≤ 2 cm, T2 is tumor \> 2 cm but ≤ 5 cm, T3 is
tumor \> 5cm, and T4 is a tumor that is any size which has directly
invaded the chest wall, skin, or is classified as inflammatory. N stage
is the number of lymph nodes involved where N1 is 1-3 positive lymph
nodes, N2 is 4-9 positive lymph nodes, and N3 is 10+ positive nodes or
specific node groups. 6th stage is the overall stage of cancer which
includes IIA, IIB, IIC, IIIA, IIIB, and IIIC. Next is the grade of the
breast cancer which is broken down into grade I, grade II, grade III,
and grade IV. A stage is defined by either regional or distant. Tumor
size is the size of the tumor at the time of diagnosis. Estrogen status
is status of the estrogen receptor which is defined as either negative
or positive. Progesterone status is the status of the progesterone
receptor which is again either negative or positive. Regional node
examined is whether the regional lymph nodes were examined and if so
then how many. Regional node positive states how many of the regional
lymph nodes were positive for cancer. Survival months states the number
of months the patient survived after diagnosis. Lastly, status states
whether the patient is alive or passed away [@Teng2019SEER].

Using this data, the analysis applies LASSO regression to perform
variable selection on the defined variables and identify the key
predictors of breast cancer survival.

ChatGPT was used to help review grammar and clarity of this
introduction.

## **Article Reviews:**

**High-Dimensional LASSO-Based Computational Regression Models**

This article summarizes high-dimensional LASSO-based computational
regression models as an effective solution for situations in which the
number of predictors is large relative to the sample size. The defining
feature of LASSO is the use of an L1 penalty that performs coefficient
shrinkage and variable selection which reduces weak or irrelevant
predictors to a coefficient of zero. This regression model is good for
biomedical, genomic, engineering applications where multicollinearity
and overfitting are common challenges. It highlights how LASSO improves
model interpretability, numerical stability, and predictive performance
compared to unregularized regression by controlling variance without
sacrificing key signal variables. Overall, it is shown that LASSO is
used for high-dimensional regression, offering a balance between model
complexity and predictive accuracy.

**LASSO Regression**

Although a short one page summary of LASSO regression, it summarizes
when how this model works and advantages of using it. The model
minimizes prediction error by putting constraints on the model
parameters which forces coefficients towards zero. Then, the
coefficients that are zero are removed from the model. A benefit of
doing this is that it reduces overfitting in the model. A disadvantage
of the model is that because the focus in on the best combined
prediction, it may not be reliable in terms of independent risk factors.

**The Generalized LASSO**

This article summarizes what the least absolute shrinkage and selection
operator (LASSO) regression model is. It describes that the function is
good for working with data that has outliers. It automatically selects
relevant variables and shrinks many of the other ones to zero. It can be
optimized to guarantee convergence which makes it good to use for
large-scale regression problems.

**Comparing the Prediction Accuracy of Ridge, LASSO and Elastic Net
Regression Models with Linear Regression Using Breast Cancer Data**

This paper used 3 different methods to determine factors that can help
determine the survivability of breast cancer patients. This is important
because it can help determine what factors can predict the survivability
of breast cancer. LASSO had the lowest prediction error and most
effective variable selection. Of the 14 variables examined, 8 of them
were significant in predicting breast cancer. Patients in the age range
of 60-90, that are in stage 2 and stage 3 of the disease have a lower
survival time. One of the limitations in this paper is that it had a
smaller number of patients that it used in Zaria-Nigeria so it was
recommended to look at a broader range of data.  

**Identification of genes associated with survival of breast cancer\
patients**

This study looked at different microRNA and mRNA data to predict
survivability in breast cancer. LASSO regression was used to identify 10
out of 56 microRNA's with non-zero coefficients that are associated with
prognosis of breast cancer patients. This is important because it can
help improve risk stratification, prognosis, and personalized treatment
planning for breast cancer patients. LASSO regression was the best
method for working with this data due to it being high-dimensional and
noisy. It was solved by classifying patients into high and low risk
survival groups. This research also included a single limited public
dataset, so additional biological experiments should be done. 

**Development of a prediction model based on LASSO regression to
evaluate the risk of non‑sentinel lymph node metastasis in Chinese
breast cancer patients with 1–2 positive sentinel lymph nodes**

This paper is similar to what we are researching although it's strictly
looking at lymph nodes as a factor in breast cancer patients. This paper
is important because it is specifically looking at Chinese patients, so
it covers Western populations and how care may be limited for Eastern
patients. LASSO regression was applied to select the most informative
predictors and build the risk score model. The model performance was
evaluated by looking at the area under the receiver operating
characteristic curve (AUC). LASSO performed well and found that 30% of
patients could possibly avoid axillary lymph node dissection
(ALND). This study was conducted at one research center so it is
recommended that this study be applied to a larger group to confirm
generalizability. 

**Comparison of lasso logistic regression, artificial neural networks,
and support vector machine in predicting breast cancer**

This study evaluates LASSO Logistic Regression as a tool for classifying
breast tumors as benign or malignant and compares its performance
against other regression models using the Wisconsin Diagnostic Breast
Cancer (WDBC) dataset. LASSO was used by adding the L1 penalty to the
function to find the strongest variables. This is shown to work well in
high-dimensional medical datasets where the data can be highly
correlated. The data was split into 80% training and 20% testing sets.
LASSO outpreformed all of the other logistic regression models that it
was compared to.

## Methods

-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.

LASSO regression can be defined by [@emmert2019high]:

$$
\hat{\beta} = arg \space \min \left\{\frac{1}{2n} \sum_{i=1}^{n} (y_i - \sum_{j} \beta_j x_{ij})^2 + \lambda||\beta||_1\right\} 
$$

where (define variables)

## Analysis and Results

### Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

A study was conducted to determine how...

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{python}
import pandas as pd
```

```{r, warning=FALSE, echo=TRUE}
# Load Data
kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

### Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
