---
title: "LASSO Regression - Spring 2026 "
subtitle: "Determining Predictors in Survival for Breast Cancer Patients"
author: "Stephanie Hopkins & Aminata Gado Oumarou (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)

## Introduction

The LASSO (Least Absolute Shrinkage and Selection Operator) regression
model has become increasingly popular over the past 30 years. It was
first introduced by Robert Tibshirani as a method for variable selection
and regularization to improve prediction accuracy and interpretability.
Since then, LASSO has been widely applied across many fields, including
biology, computational social sciences, social media, the stock market,
economics, political science, robotics, climatology, and pharmacology
[@emmert2019high].

Regression is a fundamental task in supervised machine learning, used to
model relationships between variables and address key challenges such as
parameter estimation and variable selection [@muthukrishnan2016lasso].
As machine learning methods continue to grow in popularity and show
strong results, their importance in research has increased. In medicine
especially, predictive models developed through machine learning can
help professionals distinguish between risk groups, design personalized
treatment plans, and improve patient survival outcomes
[@huang2023practical].

LASSO regression works by minimizing prediction error while applying a
constraint that shrinks some regression coefficients toward zero
[@ranstam2018lasso]. It does this by adding an L1 penalty term to the
loss function, which constrains the sum of the absolute values of the
coefficients [@ranstam2018lasso]. This penalty forces weaker predictors
to exactly zero, effectively performing automatic variable selection
[@usman2021comparing]. In other words, LASSO minimizes the residual sum
of squares subject to a constraint on the absolute values of the
coefficients, producing models that are simpler and easier to interpret
[@tibshirani1996regression]. By reducing overfitting and focusing on the
most relevant predictors, LASSO produces more interpretable models,
which is especially important in clinical survival analysis where
clarity and reproducibility matter [@ranstam2018lasso]. It is also
effective for handling large-scale problems and high-dimensional data
[@roth2004generalized].

The goal of this analysis is to use LASSO regression to identify the
factors most strongly associated with survival in breast cancer
patients. Cancer continues to have a major impact worldwide. In 2025, an
estimated 2,041,910 new cancer cases will be diagnosed in the United
States, and approximately 618,120 people are expected to die from the
disease [@nci2025statistics]. Breast cancer remains a significant public
health concern. By 2019, more than 3.8 million individuals in the United
States were living with breast cancer, and it is still the most commonly
diagnosed cancer among women and the second most common cancer globally
[@usman2021comparing]. Breast cancer begins in the breast’s feeding
tissues and develops in the epithelial cells of the ducts (85% of cases)
and lobules (15%) and is influenced by factors including hereditary
factors, age, postmenopausal obesity, alcohol or smoking, and other
factors [@hadba2024comparison]. Identifying the clinical and demographic
factors most strongly linked to survival can help improve prognostic
models and support better clinical decision-making. Previous research
has identified numerous factors associated with breast cancer prognosis
and survival. These factors are commonly grouped into three categories:
(1) demographic and genetic characteristics, such as age at diagnosis,
marital status, and reproductive history; (2) clinicopathological
factors, including tumor location, size, and histologic grade; and (3)
treatment-related variables, such as drug therapy, radiotherapy, and
chemotherapy [@teng2022dynamic].

To accomplish this, the analysis uses the Surveillance, Epidemiology,
and End Results (SEER) dataset to determine the most important variables
associated with breast cancer survivability. SEER has collected and
maintained high-quality, validated data on mortality among cancer
survivors, allowing for analysis of overall and cause-specific death
patterns [@xu2022lasso]. The dataset includes variables such as age,
race, marital status, T stage, N stage, 6th stage, grade, A stage, tumor
size, estrogen status, progesterone status, regional nodes examined,
regional nodes positive, survival months, and status [@Teng2019SEER].

Although LASSO regression has been valuable across many research areas,
it is particularly useful in medical studies where identifying the most
influential predictors is essential for improving patient outcomes.
Using the SEER data, this study applies LASSO regression for feature
selection to identify key clinical variables associated with breast
cancer survival outcomes while reducing overfitting and improving model
stability [@lee2025machine].

## Methods

-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.

LASSO regression can be defined by [@emmert2019high]:

$$
\hat{\beta} = arg \space \min \left\{\frac{1}{2n} \sum_{i=1}^{n} (y_i - \sum_{j} \beta_j x_{ij})^2 + \lambda||\beta||_1\right\} 
$$

where (define variables)

## Analysis and Results

### Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

A study was conducted to determine how...

```{r}
```

### Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
