---
title: "LASSO Regression - Spring 2026 "
subtitle: "Determining Predictors in Survival for Breast Cancer Patients"
author: "Stephanie Hopkins & Aminata Gado Oumarou (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)

## Introduction

The LASSO (Least Absolute Shrinkage and Selection Operator) regression
model has become increasingly popular over the past 30 years. It was
first introduced by Robert Tibshirani as a method for variable selection
and regularization to improve prediction accuracy and interpretability.
Since then, LASSO has been widely applied across many fields, including
biology, computational social sciences, social media, the stock market,
economics, political science, robotics, climatology, and pharmacology
[@emmert2019high].

Regression is a fundamental task in supervised machine learning, used to
model relationships between variables and address key challenges such as
parameter estimation and variable selection [@muthukrishnan2016lasso].
As machine learning methods continue to grow in popularity and show
strong results, their importance in research has increased. In medicine
especially, predictive models developed through machine learning can
help professionals distinguish between risk groups, design personalized
treatment plans, and improve patient survival outcomes
[@huang2023practical].

LASSO regression works by minimizing prediction error while applying a
constraint that shrinks some regression coefficients toward zero
[@ranstam2018lasso]. It does this by adding an L1 penalty term to the
loss function, which constrains the sum of the absolute values of the
coefficients [@ranstam2018lasso]. This penalty forces weaker predictors
to exactly zero, effectively performing automatic variable selection
[@usman2021comparing]. In other words, LASSO minimizes the residual sum
of squares subject to a constraint on the absolute values of the
coefficients, producing models that are simpler and easier to interpret
[@tibshirani1996regression]. By reducing overfitting and focusing on the
most relevant predictors, LASSO produces more interpretable models,
which is especially important in clinical survival analysis where
clarity and reproducibility matter [@ranstam2018lasso]. It is also
effective for handling large-scale problems and high-dimensional data
[@roth2004generalized].

The goal of this analysis is to use LASSO regression to identify the
factors most strongly associated with survival in breast cancer
patients. Cancer continues to have a major impact worldwide. In 2025, an
estimated 2,041,910 new cancer cases will be diagnosed in the United
States, and approximately 618,120 people are expected to die from the
disease [@nci2025statistics]. Breast cancer remains a significant public
health concern. By 2019, more than 3.8 million individuals in the United
States were living with breast cancer, and it is still the most commonly
diagnosed cancer among women and the second most common cancer globally
[@usman2021comparing]. Breast cancer begins in the breast’s feeding
tissues and develops in the epithelial cells of the ducts (85% of cases)
and lobules (15%) and is influenced by factors including hereditary
factors, age, postmenopausal obesity, alcohol or smoking, and other
factors [@hadba2024comparison]. Identifying the clinical and demographic
factors most strongly linked to survival can help improve prognostic
models and support better clinical decision-making. Previous research
has identified numerous factors associated with breast cancer prognosis
and survival. These factors are commonly grouped into three categories:
(1) demographic and genetic characteristics, such as age at diagnosis,
marital status, and reproductive history; (2) clinicopathological
factors, including tumor location, size, and histologic grade; and (3)
treatment-related variables, such as drug therapy, radiotherapy, and
chemotherapy [@teng2022dynamic].

To accomplish this, the analysis uses the Surveillance, Epidemiology,
and End Results (SEER) dataset to determine the most important variables
associated with breast cancer survivability. SEER has collected and
maintained high-quality, validated data on mortality among cancer
survivors, allowing for analysis of overall and cause-specific death
patterns [@xu2022lasso]. The dataset includes variables such as age,
race, marital status, T stage, N stage, 6th stage, grade, A stage, tumor
size, estrogen status, progesterone status, regional nodes examined,
regional nodes positive, survival months, and status [@sembay2021seer].

Although LASSO regression has been valuable across many research areas,
it is particularly useful in medical studies where identifying the most
influential predictors is essential for improving patient outcomes.
Using the SEER data, this study applies LASSO regression for feature
selection to identify key clinical variables associated with breast
cancer survival outcomes while reducing overfitting and improving model
stability [@lee2025machine].

## Methods

-   Detail the models or algorithms used.

-    Justify your choices based on the problem and data.

## Analysis and Results

### Data Exploration and Visualization

This study uses breast cancer data from the Surveillance, Epidemiology,
and End Results (SEER) program, which is maintained by the National
Cancer Institute (NCI). The dataset, composed of 4,024 patients,
includes demographic, clinical, and outcome variables for female
patients diagnosed with infiltrating ductal and lobular breast carcinoma
between 2006 and 2019. These variables are well suited for predictive
modeling of patient survival outcomes. The dataset excludes patients
with unknown tumor size, examined regional lymph nodes (LNs), regional
positive LNs, and patients whose survival months were less than 1 month
[@sembay2021seer].

| Category | Variables |
|------------------------------------|------------------------------------|
| Demographics | Age at diagnosis, race, marital status (married, single, widowed, divorced, separated) |
| Tumor Characteristics | T stage (T1–T4), N stage (N1–N3), 6th AJCC stage (IIA–IIIC), tumor grade (I–IV), A stage (regional/distant), tumor size |
| Hormone Receptors | Estrogen status, progesterone status (positive/negative) |
| Lymph Nodes | Nodes examined, nodes positive |
| Outcome | Survival months, vital status (alive/dead) |

: Table 1. Characteristics of the variables included in the study.

Using the variables defined above, analysis of the variables to predict
survival will be analyzed. To start, below is the distribution of age at
the time of breast cancer diagnosis.

```{r}
library(ggplot2)
data <- read.csv("SEER_Breast_Cancer_Data.csv")
```

```{r}
ggplot(data, aes(x = Age)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "black", alpha = 0.7) +
  labs(title = "Age at Diagnosis",
       x = "Age (Years)",
       y = "Frequency") +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

```

This histogram shows that patients are between ages 30 to 70 with a
majority of diagnoses aged 50 and older. Due to the data that has been
excluded in this study, the age of the women that have been diagnosed
with breast cancer could vary if it were to be included. The next
analysis shows a boxplot of survival based on stages of breast cancer.

```{r}
ggplot(data, aes(x = X6th.Stage, y = Survival.Months)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Survival by AJCC Stage",
       x = "6th AJCC Stage",
       y = "Survival Months") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) 

```

The boxplot shows that in general there is a general downward trend in
survival months based on 6th AJCC cancer stages (IIA, IIB, IIIA, IIIB,
IIIC). The early stages of breast cancer (IIA, IIB, IIIA) show a higher
median survival where later stages (especially IIIC) show a lower median
survival. This is suggesting that a higher stage breast cancer diagnosis
is associated with shorter survival. Although, the median survival
across the stages do not differ dramatically which shows that there is
overlap between groups, thus stage alone may not fully explain a
survival outcome. Due to the length of the whiskers, the length of
survival varies widely within each stage and there may be other
predictors that likely influence survival.

```{r}
#install.packages("survminer")
library(survminer)
library(survival)

data$event <- ifelse(data$Status == "Dead", 1, 0)

fit <- survfit(Surv(Survival.Months, event) ~ 1, data = data)

plot <- ggsurvplot(fit,
                   data = data,
                   conf.int = TRUE,
                   risk.table = TRUE,
                   pval = FALSE,
                   xlab = "Time (Months)",
                   ylab = "Survival Probability",
                   title = "Kaplan–Meier Survival Curve",
                   palette = "blue")

plot$plot <- plot$plot +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

plot


```

A Kaplan–Meier survival analysis was conducted to estimate the overall
survival over a 100-month follow-up period. The survival curve indicates
a gradual decline in survival probability over time, beginning at 100%
at baseline and decreasing steadily to approximately 80% by 100 months.
The stepwise pattern reflects the occurrence of events throughout the
study period, while censoring is indicated by tick marks along the
curve. The number at risk decreased over time, from 4,024 participants
at baseline to 444 at 100 months, reflecting both events and censoring.
Overall, the results suggest relatively high survival across the study
population with a slow and consistent reduction in survival probability
over time.

### **Statistical Analysis**

The selected demographic, clinical, and outcome variables were used as
predictors in a LASSO regression model to identify factors associated
with patient survival.

The data was first prepared in order to be used for LASSO regression.
Categorical predictors were converted into indicator variables so they
could be included in the model. Continuous variables were scaled to
common ranges to ensure that no single variable is dominated by the
penalty applied by LASSO. Specifically, a penalized Cox proportional
hazards model will be used for censored survival data. Observations that
had missing values were removed to maintain consistency in the
analysis.  

Variable selection was performed using penalized Cox regression with an
L1 (LASSO) penalty. LASSO regression can be defined by
[@emmert2019high]:

$$
\hat{\beta} = arg \space \min \left\{\frac{1}{2n} \sum_{i=1}^{n} (y_i - \sum_{j} \beta_j x_{ij})^2 + \lambda||\beta||_1\right\} 
$$

where $y_{i}$ represents the observed outcome, $x_{ij}$ denotes
predictor values, β is the vector of regression coefficients, and λ is a
tuning parameter which controls the strength of the penalty. The L1
penalty encourages sparsity by shrinking weaker coefficients to zero
which is what performs the automatic variable selection. 

The analysis was conducted in R using the **glmnet** package which is
designed for fitting penalized regression models. The strength of the
penalty is controlled by λ. To select an appropriate value of λ, k-fold
cross-validation was used. This process constantly splits the data into
training and testing sets to evaluate the model performance and choose
the penalty that produces the best balance between simplicity and
predictive accuracy. 

Predictors with non-zero coefficients in the final LASSO model were
interpreted as significant variables associated with breast cancer
survival. 

## Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

## Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
