---
title: "Determining Predictors in Survival for Breast Cancer Patients Using LASSO Regression"
subtitle: ""
author: "Stephanie Hopkins & Aminata Gado Oumarou (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  revealjs:
    slide-number: true
    theme: default
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

## Introduction: What is LASSO Regression?

::: {style="font-size: 24px;"}
-   LASSO - Least Absolute Shrinkage and Selection Operator

-   Introduced by Robert Tibshirani in 1996

-   Method for variable selection and regularization to improve
    prediction accuracy and interpretability [@ranstam2018lasso]

    -   shrinks less important variables to zero

    -   helps prevent overfitting

-   Widely used - medicine, biology, computational social sciences,
    social media, the stock market, economics, political science,
    robotics, climatology, and pharmacology [@emmert2019high]
:::


## Introduction: Breast Cancer

::: {style="font-size: 24px;"}

- In 2019, over 3.8 million people in the U.S. were living with breast cancer. In 2025, 2.04 million new cancer cases were diagnosed. It is the most commonly diagnosed cancer among women and the second most common cancer globally [@usman2021comparing] [@nci2025statistics]

- There are many causes, including hereditary factors, age, obesity, alcohol or smoking, and other factors

- Prognosis and survival are commonly grouped into three categories: (1) demographic and genetic characteristics, (2) clinicopathological factors, and (3) treatment-related variables such as drug therapy, radiotherapy, and chemotherapy [@teng2022dynamic]

:::

:::: {.columns}

::: {.column width="50%"}

```{r, echo=FALSE}
library(glmnet)
library(survminer)
library(survival)
library(ggplot2)
library(corrplot)

seer_data <- read.csv("SEER_Breast_Cancer_Data.csv")

ggplot(seer_data, aes(x = Age)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "black", alpha = 0.7) +
  labs(
    title = "Age at Diagnosis",
    x = "Age (Years)",
    y = "Frequency"
  ) +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```
:::

::: {.column width="50%"}
```{r}
ggplot(seer_data, aes(x = Status, y = Tumor.Size, fill = Status)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3) +
  labs(
    title = "Tumor Size by Survival Status",
    x = "Survival Status",
    y = "Tumor Size"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```
:::

::::



## Methods
::: {.panel-tabset}

### LASSO Regression
::: {style="font-size: 20px;"}

-   LASSO is defined as:

$$ \hat{\beta} = arg \space \min \left\{\frac{1}{2n} \sum_{i=1}^{n} (y_i - \sum_{j} \beta_j x_{ij})^2 + \lambda||\beta||_1\right\}  $$

-   $y_{i}$ represents the observed outcome

-   $x_{ij}$ denotes predictor values

-   β is the vector of regression coefficients

-   λ is a tuning parameter

-   L1 penalty encourages sparsity by shrinking weaker coefficients
    exactly to zero
:::

---

### LASSO-Cox Regression
::: {style="font-size: 20px;"}


-   For the LASSO-Cox model, coefficients are instead estimated by
    maximizing the penalized Cox partial likelihood:

$$ \hat{\beta}_{\text{LASSO-Cox}} = \arg_{\beta}\max\left \{PL(\beta) - \lambda \sum_{j=1}^{p}|\beta_j| \right\} $$

-   PL(β) denotes the Cox partial likelihood function 

-   λ is a positive tuning parameter that controls the amount of
    shrinkage

-   L1 penalty shrinks some regression coefficients to zero while
    estimating hazard ratios for the remaining predictors

**Assumptions:**

-   Hazard ratios between individuals remain constant over time
    (proportional hazards assumption)

-   Survival times of different patients are independent from one
    another

-   Censoring is assumed to be non-informative, meaning that the reason
    for censoring is unrelated to the probability of the event
:::
:::

## Analysis and Results

::: {style="font-size: 20px;"}
-   Determine predictors in survival for breast cancer patients using
    Surveillance, Epidemiology, and End Results (SEER) dataset

-   Composed of 4,024 patients with ductal and lobular breast carcinoma
    between 2006 and 2019

-   Excludes patients with unknown tumor size, examined regional lymph
    nodes (LNs), regional positive LNs, and patients whose survival
    months were less than 1 month

| Category | Variables |
|------------------------------------|------------------------------------|
| Demographics | Age at diagnosis, race, marital status (married, single, widowed, divorced, separated) |
| Tumor Characteristics | T stage (T1–T4), N stage (N1–N3), 6th AJCC stage (IIA–IIIC), tumor grade (I–IV), A stage (regional/distant), tumor size |
| Hormone Receptors | Estrogen status, progesterone status (positive/negative) |
| Lymph Nodes | Nodes examined, nodes positive |
| Outcome | Survival months, vital status (alive/dead) |

: Table 1. Characteristics of the variables included in the study.
:::

## Kaplan-Meier Estimate of Survival

::: {style="font-size: 24px;"}
```{r}
seer_data$event <- ifelse(seer_data$Status == "Dead", 1, 0)

fit <- survfit(Surv(Survival.Months, event) ~ 1, data = seer_data)

plot <- ggsurvplot(fit,
                   data = seer_data,
                   conf.int = TRUE,
                   censor = FALSE,
                   risk.table = TRUE,
                   size = 1.2, 
                   pval = FALSE,
                   xlab = "Time (Months)",
                   ylab = "Survival Probability",
                   title = "Kaplan–Meier Survival Curve",
                   palette = "blue")

plot$plot <- plot$plot +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

plot


```

-   Survival curve indicates a gradual decline in survival probability
    over time, beginning at 100% at baseline and decreasing steadily to
    approximately 78% by 100 months

-   Relatively high survival across the study population with a slow and
    consistent reduction in survival probability over time
:::

## Kaplan-Meier Survival by 6th Stage

::: {style="font-size: 20px;"}
```{r}
seer_data$event <- ifelse(seer_data$Status == "Dead", 1, 0)
seer_data$X6th.Stage <- as.factor(seer_data$X6th.Stage)

fit <- survfit(Surv(Survival.Months, event) ~ X6th.Stage, data = seer_data)

plot <- ggsurvplot(fit, 
                   data = seer_data, 
                   conf.int = TRUE, 
                   risk.table = TRUE, 
                   risk.table.height = .40, 
                   risk.table.fontsize = 3.5, 
                   pval = FALSE, 
                   xlab = "Time (Months)",
                   ylab = "Survival Probability",
                   title = "Kaplan–Meier Survival Curves by 6th Stage",
                   legend.labs = c("IIA","IIB","IIIA","IIIB","IIIC"))
plot$plot <- plot$plot + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

plot

```

-   Patients diagnosed at earlier stages (IIA, IIB) generally show
    higher survival probabilities throughout the follow-up period

-   Patients diagnosed at later stages (IIIA, IIIB, IIIC) experience
    a faster decline in survival probability over time

-   Suggests patients diagnosed with more advanced stages tend to have
    lower survival probabilities compared to those diagnosed at earlier
    stages
:::




## Modeling and Results

::: {style="font-size: 24px;"}
-   LASSO-penalized Cox proportional hazards model was used to identify
    key predictor of breast cancer survival

-   Categorical predictors were converted into indicator (dummy)
    variables

-   Continuous variables were standardized automatically within the
    glmnet framework to ensure that no single variable dominated the L1
    penalty applied by LASSO

-   Observations with missing values were removed to maintain
    consistency in the analysis.

```{r}
#remove observations with missing values
vars <- c("Age", "Race", "Marital.Status", "T.Stage", "N.Stage", "X6th.Stage", "Grade", "A.Stage", "Tumor.Size", "Estrogen.Status", "Progesterone.Status", "Regional.Node.Examined", "Reginol.Node.Positive", "Survival.Months", "Status")

#Subset the dataset to include only the variables used in the survival analysis

model_data <- seer_data[, vars]

#Remove observations with missing values
model_data <- na.omit(model_data)

#Create an event indicator variable required for survival analysis
# 1 indicates the event occurred (death) and 0 indicates censored observations (alive)
model_data$event <- ifelse(model_data$Status == "Dead", 1, 0)

```

-   Survival object was created using the survival time and event
    indicator

-   Predictor variables were converted into a model matrix

-   Fit model using glmnet which estimated regression coefficients while
    applying regularization to perform variable selection
:::

## Coefficient Path Plot

::: {style="font-size: 20px;"}
```{r}
y <- Surv(model_data$Survival.Months, model_data$event)

#x <- model.matrix(Surv(Survival.Months, event) ~ Age + Race + Marital.Status + T.Stage + N.Stage + X6th.Stage + Grade + A.Stage + Tumor.Size + Estrogen.Status + Progesterone.Status + Regional.Node.Examined + Reginol.Node.Positive, data = model_data)[,-1]

x <- model.matrix(event ~ . - event - Status - Survival.Months, data = model_data)[,-1]


#Fit a penalized cox regression model using glmnet
fit <- glmnet(x, y, family = "cox")
plot(fit)
```

-   Demonstrates how predictor effects evolve as the penalty parameter
    (λ) decreases

-   The x-axis, -log(λ), represents the strength of the penalty

-   The y-axis displays the estimated coefficients, indicating the
    magnitude and direction of each predictor's effects

-   The numbers along the top of the model represent the number of
    non-zero coefficients at each value of λ
:::

## Optimal Lambda Value

::: {style="font-size: 20px;"}
```{r}
#Perform cross-validation to identify the optimal lambda value
cvfit <- cv.glmnet(x, y, family = "cox", alpha = 1)

#Plot the cross-validation error across lambda values
plot(cvfit)
```

-   Displays the mean cross-validation error and associated confidence
    intervals for each λ

-   Error decreases as more variables are included, then stabilizes

-   Optimal λ is selected using cross-validation

-   Indicates that a simpler model performs similarly to a more complex one
:::

## LASSO-Cox Model Results

::: {style="font-size: 15px;"}
```{r}
library(knitr)
lasso_coef <-coef(cvfit, s = "lambda.min")
coef_vector <- as.vector(lasso_coef)

#Build data frame
coef_df <- data.frame(
  Variable = rownames(lasso_coef),
  Coefficient = coef_vector,
  stringsAsFactors = FALSE
)

#Keep only non-zero coefficients
coef_df <- coef_df[!is.na(coef_df$Coefficient), ]
coef_df <- coef_df[coef_df$Coefficient !=0, ]

#Add interpretation
coef_df$Interpretation <- ifelse(
  coef_df$Coefficient > 0,
  "Increased hazard (worse survival)",
  "Decreased hazard (better survival)"
)

#Print table
kable(
  coef_df,
  caption = "Selected Predictors from the LASSO-Cox Model",
  digits = 3,
  align = "lcc"
)
```
:::

## Conclusion

::: {style="font-size: 24px;"}
-   Identified 22 predictors with non-zero coefficients at the optimal
    penalty parameter

-   Positive coefficients linked to an increased hazard, or worse
    survival

-   Negative coefficients linked to a decreased hazard, or better
    survival

-   Five most influential variables selected: Grade IV
    (Undifferentiated; Anaplastic) ($\beta = 1.058$), Other races
    (American Indian/AK Native, Asian/Pacific Islander)
    ($\beta = -0.649$), N stage N3 ($\beta = 0.648$), estrogen receptor
    positive status ($\beta = -0.641$), and T stage T4 ($\beta = 0.599$)

-   Patients with Grade IV tumors, N stage N3, and T stage T4 exhibited
    worse survival outcomes

-   Other races and those with positive estrogen receptor status showed
    improved survival

-   Both demographic and biological factors play a meaningful role in
    prognosis

-   Different penalty parameter (λ) could lead to a slightly different
    set of predictors
:::

## References
